---
title: "Assignment 2: Coding Basics"
author: "Nancy Bao"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
seqbyfour<-seq(1,100,4) # I used the seq(start#, to#, by#) function to generate a sequence of numbers starting from 1 and ending at 100, and specified with increments of four. 
seqbyfour #I named the sequence "seqbyfour" and calling it in this line.
#2. 
seqbyfour_mean<-mean(seqbyfour) # I used mean() of the sequence, "seqbyfour". I created an object for the mean and specified the name of the value as seqbyfour_mean.
seqbyfour_mean #calling the mean in this line. The mean of the seqbyfour sequence is 49. 
seqbyfour_median<-median(seqbyfour) # I used median() on "seqbyfour" to find the median. I an object for the median and specified name of the median value as seqbyfour_median. 
seqbyfour_median #calling the median in this line. The median is 49 for this sequence. 
#3. 
seqbyfour_mean>seqbyfour_median #I used the conditional statement on the objects I created to determine if the mean was greater than the median of the seqbyfour sequence. The answer returned is FALSE. So, the mean is not greater than the median. We can confirm this as the mean value = median value =49. 
#4 The comments (preceded by `#`) detailing reasoning/steps taken, are annotated after the code portion in each question.   
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
Student_names<-c("Robert","Luna","Emily","Zachary")
Test_scores_outof100<-c(98,48,85,79)
test_scores_pass50<-c(TRUE,FALSE,TRUE,TRUE)
#Combine the vectors into a data frame
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
